---
title: "individual2.Rmd"
author: "Mike Brown"
date: "3/4/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
install.packages('readxl')
dat <- readxl::read_xls('Users/brown/OneDrive/Documents/ds202_lab3/GSS.xls')
``` 

1.Find the average years of education of he respondents with each marital status. Arrange your output in a meaningful order and print. Describe any patterns you find. 
``` {r} 
a<- dat$`Marital status` %>% factor
b<- as.numeric(dat$`Highest year of school completed`)

ggplot(dat, aes(x=a, y=b))+geom_boxplot()

```

  Answer: From this boxplot we can see that there is a trend that seperated and widowed individuals tend to have lower amounts education.

2.Create a single data frame containing records for the better educated respondents with each marital status. A "better educated" respondent is someone who has strictly more years of education than the average among those with the same marital status. Print the structure of the data frame. 

  Answer:
  
**For each of the investigation below, create at least one nemerical summary and one appropriate visualization. Write a paragraph to describe your findings and the variables you used.**

3.How is the happiness of a respondent related to his/her marriage status? Define that a person is happy if the resonse to question is "Very happy" or "Pretty happy".

  Answer: Married individuals appear to have a much greater presence in the happy categories while the other marital statuses didn't display much of a pattern anywhere else. Certainly married individuals are more likely to be happy than others. 

4.Does party affiliation affect the relationship you found in the last question?

  Answer: Party affiliation doesn't appear to have an effect on marital status. 

5.How is marital status related to the education of a respondent? What are possible reasons accounting for the patterns you see?

  Answer: Divorced and never married individuals have a higher average amount of education than the rest, while 20 years of education is not considered an outlier or uncommon for married and never married individuals. 
  
6.Explore two more interesting questions, and answer those questions using the GSS data. 

  Answer: happiness of people before 2000 v happiness of people after 2000
  
  
  
  number of hours spouse works a week before 2000 v after 2000
  
  